Statistical Reporting Inconsistencies in Experimental Linguistics

### 2024-07-18 manuscript/manuscript.tex
edits and added conclusion

\title{Statistical Inconsistencies in Experimental Linguistics}

@@

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

What we know about human language and its cognitive underpinnings is
often informed by experimental data, data based on which researchers
draw theoretically relevant inferences using common statistical
frameworks. This inference should be transparent in order for other
researchers to critically evaluate the inference process and potentially
detect and correct human errors. In the recent years, the quantitative
sciences have seen repeated calls to become more transparent and
reproducible by sharing data and statistical protocols
(\citeproc{ref-arvan2022reproducibility}{Arvan et al., 2022};
\citeproc{ref-laurinavichyute2022share}{Laurinavichyute et al., 2022};
\citeproc{ref-roettger2019researcher}{Roettger, 2019}), sharing of
statistical protocols is still rare across the language sciences
(\citeproc{ref-bochynska2023reproducible}{Bochynska et al., 2023}). If
statistical procedures cannot be critically evaluated, human errors
might be left undetected and thus remain uncorrected in the publication
record. If these errors affect the decision procedure of the analysis,
i.e.~whether a hypothesis is rejected or accepted, these errors might
lead to -at best- overconfident, -at worst- false theoretical
conclusions. The present paper will present evidence that the published
literature in experimental linguistics contains a concerning amount of
statistical errors, a state of affairs which warrants more rigorous data
sharing practices.

@@

The null-hypothesis significance testing (NHST) framework is, to date,
the most dominant statistical framework that researchers use to test
hypotheses in the language sciences
(\citeproc{ref-sondereggera2024advancements}{Sonderegger \& Sóskuthy,
2024}). These statistical tests are reported in specific formats which
usually contain a test statistic, the degrees of freedom of that test
(if applicable), and the p-value, representing the probability of
observing the data (or more extreme data) given the null hypothesis
(i.e.~given that the test statistic is zero) (see example 1):\\
(1) F(1, 66) = 3.88, p \textless{} .05\\
Since data and statistical protocol sharing remains rare across the
language sciences (\citeproc{ref-bochynska2023reproducible}{Bochynska et
al., 2023}), interested readers are left with trusting the authors that
the statistical analysis is run and reported correctly. Whether that
trust is warranted can be assessed, however. The three sets of indices
in (1) have a clearly defined mathematical relationship and can thus be
checked for consistency. An F test with the specified degrees of freedom
and a test-statistic of 3.88 should result in a p-value of 0.053 which
is larger, not smaller, than 0.05. Thus, something is wrong with the
statistical report in (1). Possible reasons for this inconsistency are
manifold: It could be a typo of the comparison sign, i.e.~the authors
meant to use \texttt{=} or \texttt{\textgreater{}} rather than
\texttt{\textless{}}. Alternatively, any of the numbers could be a typo.
Sometimes an error might indicate erroneous rounding (e.g.~0.057 being
rounded down to 0.05). Other times, human error along the data
analytical pipeline might have caused an error down the line. Without
access to data and scripts, it remains unclear to the reader, what has
caused the inconsistency. Such inconsistencies can be particularly
concerning if the calculated p-value and the reported p-values are not
on the same side of the alpha threshold. In NHST, p-values below a
conventionalized alpha threshold, most commonly 0.05, are interpreted as
evidence that the data are sufficiently inconsistent with the null
hypothesis (significant). p-values above that threshold are considered
consistent with the null hypothesis and practically lead to rejecting
the alternative hypothesis (not significant). In (1) above, the reported
p-value suggest a significant result, the p-value derived from the
degrees of freedom and the test statistic suggest a non-significant
result. In the following we refer to these inconsistencies as ``decision
inconsistency''.

This form of inconsistency assessment can be automatically assessed if
statistical tests are reported in an unambiguous format. Recently, a
series of studies used such automatic assessments to evaluate the
prevalence of inconsistent statistical reporting across different
disciplines (\citeproc{ref-buckley2023estimating}{Buckley et al., 2023};
\citeproc{ref-green2018statcheck}{Green et al., 2018};
\citeproc{ref-gross2021fidelity}{Groß, 2021}; e.g.
\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016};
\citeproc{ref-nuijten2020statcheck}{Nuijten \& Polanin, 2020}). For
example, looking at over 250000 p-values published in major psychology
journals, Nuijten et al. (\citeproc{ref-nuijten2016prevalence}{2016})
found that around 50\% of the articles with statistical results
contained at least one inconsistencies and around 12.5\% contained at
least one ``decision inconsistency''.

To assess the prevalence of statistical-reporting inconsistencies in
experimental linguistics, the present paper conceptually replicates
Nuijten et al. (\citeproc{ref-nuijten2016prevalence}{2016}) and assesses
over 39000 p-values reported in eight experimental linguistic journals
published between 2000 and 2023. We further assess whether the rate of
inconsistency differs across journals, whether that rate has changed
over the course of the last 20 years and whether there is evidence for
bias in these statistical-reporting inconsistencies. We discuss the
results and offer concrete recommendations for authors, reviewers, and
editors to tackle this problem.

@@

All quantitative analyses were conducted using R {[}{]} and the r
packages {[}LIST ALL PACKAGES from 01\_Analysis.R and add to
references{]}

@@

We used the R package statcheck (Version 1.4.1-beta.2,
\citeproc{ref-nuijten2023statcheck}{Nuijten \& Epskamp, 2023}) to
automatically detect statistical-reporting inconsistencies. Statcheck
works as follows: After converting pdf or html articles to plain text,
statcheck searches for specific strings that correspond to a NHST
result, using ``regular expressions''. That way, statcheck can detect
results of t tests, F tests, Z tests, χ2 tests, correlation tests, and Q
tests as long as the test result fulfills three conditions: (a) the test
result is reported completely including the test statistic, degrees of
freedom (if applicable), and the p-value; (b) the test result is in the
body of the text, i.e.~it usually misses results reported in tables; and
(c) the test result is reported in American Psychological Association
style (\citeproc{ref-APA2020}{American Psychological Association,
2020}). Given these constraints, statcheck is estimated to detect
roughly 60\% of all reported NHST results
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}) Statcheck
uses the reported test statistic and degrees of freedom to recalculate
the p-value, compares the reported and recalculated p-values and, if
there is a mismatch, reports a comparison as an ``inconsistency.'' The
algorithm takes into account that tests might have been performed as
one-tailed by identifying the search strings ``one-tailed,''
``one-sided,'' or ``directional''. Moreover, statcheck counts p = .000
and p \textless{} .000 as inconsistent as p-values of exactly zero are
mathematically impossible and the APA manual
(\citeproc{ref-APA2020}{American Psychological Association, 2020})
advises to report very small p-values as p \textless{} .001. Validity
checks suggest that inter rater reliability between manual coding and
statcheck is high, i.e.~0.76 for inconsistencies and 0.89 for decision
inconsistencies (\citeproc{ref-nuijten2016prevalence}{Nuijten et al.,
2016}). The overall accuracy of statcheck is estimated between 96.2\% to
99.9\% (\citeproc{ref-nuijten2017validity}{Nuijten et al., 2017}). We
thus consider statcheck a valid, but rough proxy of statistical
reporting inconsistencies.

@@

Focusing on experimental linguistic research, we started with Kobrock
and Roettger (\citeproc{ref-kobrock2023assessing}{2023}) as a point of
departure who list 100 linguistic journals that had at least a hundred
articles published at the time of assessment (2021) and a high ratio of
articles containing the term ``experiment* in title, abstract or
keywords. Out of these 100 journals, we selected all journals with at
least 10\% of articles containing the search string''experiment*``. Out
of the remaining 37 journals, we selected those journals that urged APA
formatting either in the main body of the text or specifically regarding
statistics in the author guidelines, resulting in nine remaining
journals. Moreover, to download the pdfs, the articles had to be either
accessible to us through our library license, or open access, resulting
in a final list of eight journals: Applied Psycholinguistics (APS),
Bilingualism: Language and Cognition (BLC), Linguistic Approaches to
Bilingualism (LAB), Language and Speech (LaS), Language Learning and
Techology (LLT), Journal of Language and Social Psychology (LSP),
Journal of Child Language (JCL), and Studies in Second Language
Acquisition (SLA).

We included only original research articles within the publication years
of 2000-2023, excluding any book reviews, response articles,
commentaries, editorials, corrigenda, errata, advertisements, etc.
Articles from LAB spanned 2011-2023, while the rest spanned 2000-2023.
This procedure resulted in 5804 research articles. All 5804 articles
were submitted to analysis, using Statcheck v1.5.0. 157 articles were
not analyzable by statcheck, and thus had to be removed from the pool,
likely related to issues with rendering the Chi-Squared symbol being
erroneously converted from .pdf to .txt.

@@

All derived data and corresponding R scripts are publicly available
here: LINK. The original journal articles cannot be bulk-shared due to
distribution restrictions by the publishers. NEED TO FIGURE OUT A WAY

@@

The results are summarized in Table~\ref{tbl-summary}. Out of 5804
articles, 3059 articles contained statistical tests that statcheck could
assess (53\%), amounting to 39532 assessable p-values. 5166 p-values
were flagged as inconsistent (13.1\%) and 655 of which were considered
decisions inconsistencies (1.7\%) (see Table~\ref{tbl-summary}).

@@

{\caption{{Summary. Applied Psycholinguistics (APS), Bilingualism:
Language and Cognition (BLC), Linguistic Approaches to Bilingualism
(LAB), Language and Speech (LaS), Language Learning and Techology (LLT),
Journal of Language and Social Psychology (LSP), Journal of Child
Language (JCL), and Studies in Second Language Acquisition
(SLA).}{\label{tbl-summary}}}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0755}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1698}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1887}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1792}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1509}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2358}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Journal
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
eligible articles
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
assessible articles
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
assessible results
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
inconsistencies
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
decision inconsistencies
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
APS & 953 & 690 & 9570 & 1368 & 170 \\
BLC & 964 & 610 & 9093 & 1161 & 120 \\
JCL & 1109 & 529 & 6240 & 750 & 69 \\
LAB & 471 & 133 & 1719 & 234 & 25 \\
LLT & 421 & 111 & 919 & 201 & 61 \\
LSP & 695 & 376 & 4320 & 429 & 60 \\
LaS & 598 & 363 & 4717 & 552 & 86 \\
SLA & 593 & 247 & 2954 & 471 & 64 \\
Total & 5804 & 3059 & 39532 & 5166 & 655 \\
\end{longtable}

@@

\caption{\label{fig-stacked-bar}Proportion of articles containing at
least one inconsistency / decision inconsistency. Applied
Psycholinguistics (APS), Bilingualism: Language and Cognition (BLC),
Linguistic Approaches to Bilingualism (LAB), Language and Speech (LaS),
Language Learning and Techology (LLT), Journal of Language and Social
Psychology (LSP), Journal of Child Language (JCL), and Studies in Second
Language Acquisition (SLA).}

@@

\caption{\label{fig-year-prop}Proportion of inconsistencies / decision
inconsistencies from 2000 to 2023. Left: global rates with all journals
pooled; Right: rates per journal. Applied Psycholinguistics (APS),
Bilingualism: Language and Cognition (BLC), Linguistic Approaches to
Bilingualism (LAB), Language and Speech (LaS), Language Learning and
Techology (LLT), Journal of Language and Social Psychology (LSP),
Journal of Child Language (JCL), and Studies in Second Language
Acquisition (SLA).}

@@

When examining reported against recalculated p-values (see
Figure~\ref{fig-scatter}) for all inconsistencies, we can identify
certain spatial patterns (only inconsistencies that are based on `='
comparisons were included to make them easier to interpret, n = 5154 out
of 5166). First, the center of density of points is located in the
bottom left corner with more values reported closer to the alpha level.
This is not surprising, since there is a documented publication bias in
published quantitative articles, with hypotheses being much more often
confirmed (and thus p \textless{} .05) than not
(\citeproc{ref-franco2014publication}{Franco et al., 2014};
\citeproc{ref-sterling1959publication}{Sterling, 1959}). Second, there
is a linear density of points along the diagonal axis which corresponds
to numerically small inconsistencies, some of which might be related to
simple rounding errors. However, comparing the diagonal to the black
line, which represents a linear model predicting recalculated by
reported p-values, we can see a clear divergence of what is expected if
inconsistencies were equally likely in both directions
(i.e.~recalculated p-values were as likely to be smaller than the
reported p-value as larger). The slope of the regression line is flatter
than the diagonal axis which means that, on average, reported p-values
are lower than their recalculated counter parts. In other words,
inconsistencies have a tendency to lead to smaller p-values. This
potential bias can also be observed in decision inconsistencies. Of all
decisions inconsistencies (n = 655), 71\% represent cases in which a
reported significant result (p \textless{} 0.05) is recalculated as
non-significant value (p \textgreater{} 0.05).

@@

\subsection{4.2 Manual inspection of decision
inconsistencies}\label{manual-inspection-of-decision-inconsistencies}

To further understand the nature of decision inconsistencies, we
manually inspected all of them (n=655). We (a) evaluated whether
statcheck has extracted the information correctly, and (b) assessed
whether the text actually suggest an erroneous inferential decision,
i.e.~whether the paper uses a reported significant result to claim a
significant effect or a reported non-significant result to claim a null
result.

@@

The present study found a concerning amount of statistical reporting
inconsistencies across a sample of 5804 experimental linguistic
articles, containing 3059 assessable p-values. 13.1\% of all p-values
were flagged as inconsistent and 1.7\% were flagged as decision
inconsistencies, i.e.~the reported p-value is on the opposite side of
the alpha threshold than the recalculated p-value. On average, 50\% of
assessable articles contained at least one inconsistency and 13\%
contained at least one decision inconsistency.

>>>>

\title{Statistical Reporting Inconsistencies in Experimental
Linguistics}

@@

\usepackage{annotate-equations}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

What we know about human language and its cognitive underpinnings is
often informed by experimental data. Researchers test theoretical
predictions with these data using statistical tests. Depending on the
results of such tests, researchers make claims for or against
theoretical assumptions.\\
Since they play such an integral part in the reasoning process of
experimentalists, both the data these tests are based on and the
computational procedure of the tests itself should be transparent in
order for other researchers to critically evaluate them. Moreover things
can go wrong. To err is human. Transparent sharing allows others to
detect and correct human error. In the recent years, the quantitative
sciences have seen repeated calls to become more transparent and
reproducible by sharing data and statistical protocols
(\citeproc{ref-arvan2022reproducibility}{Arvan et al., 2022};
\citeproc{ref-laurinavichyute2022share}{Laurinavichyute et al., 2022};
\citeproc{ref-roettger2019researcher}{Roettger, 2019}). Despite these
calls, sharing of statistical protocols is still rather rare across the
language sciences (\citeproc{ref-bochynska2023reproducible}{Bochynska et
al., 2023}). If statistical procedures cannot be critically evaluated,
human errors might be left undetected and thus remain uncorrected in the
publication record. And if undetected errors affect the decision
procedure of the analysis, i.e.~whether a hypothesis is rejected or
accepted, these errors might lead to -at best- overconfident, -at worst-
false theoretical conclusions. The present paper will present evidence
that the published literature in experimental linguistics contains a
concerning amount of such statistical errors, a state of affairs which
warrants more rigorous data sharing practices.

@@

The null-hypothesis significance testing (NHST) framework is, to date,
the most dominant statistical framework that researchers use to test
hypotheses in the language sciences
(\citeproc{ref-sondereggera2024advancements}{Sonderegger \& Sóskuthy,
2024}). NHST tests are commonly reported in specific formats which
usually contain the name of the test (e.g.~F, t, χ2), a test statistic,
the degrees of freedom of that test (if applicable), and the p-value,
representing the probability of observing the data (or more extreme
data) given the null hypothesis (i.e.~given that the test statistic is
zero) (see example 1): ~

~

\begin{equation*}
  \tikzmarknode{node1}{(1) }
  \eqnmarkbox[gray]{node2}{F}
  \eqnmarkbox[orange]{node3}{(1,66)}
  \tikzmarknode{node4}{=}
  \eqnmarkbox[purple]{node5}{3.88}
  \tikzmarknode{node6}{, }
  \eqnmarkbox[blue]{node7}{p < .05}
\end{equation*}

\annotate[yshift=1em]{left}{node2}{name of test} 
\annotate[yshift=-1em]{below,left}{node3}{degrees of freedom} 
\annotate[yshift=1em]{right}{node5}{test statistic} 
\annotate[yshift=-1em]{below,right}{node7}{p-value}

~

Since data and statistical protocol sharing remains rare across the
language sciences (\citeproc{ref-bochynska2023reproducible}{Bochynska et
al., 2023}), interested readers are left with trusting the authors that
the statistical analysis is run and reported correctly. However, the
three sets of indices in (1) have clearly defined mathematical
relationships and can thus be easily checked for consistency. An F test
with the specified degrees of freedom and a test-statistic of 3.88
should result in a p-value of 0.053 which is larger, not smaller, than
0.05. Possible reasons for this inconsistency are manifold: It could be
a typo of the comparison sign, i.e.~the authors meant to use \texttt{=}
or \texttt{\textgreater{}} rather than \texttt{\textless{}}. Any of the
numbers could contain a typo. Sometimes an error might indicate
erroneous rounding (e.g.~0.057 being rounded down to 0.05). Without
access to data and scripts, it remains unclear to the reader what has
caused this inconsistency. Such inconsistencies can be particularly
concerning if the calculated p-value (here 0.057) and the reported
p-values (here \textgreater0.05) are not on the same side of the alpha
threshold. In NHST, p-values below a conventionalized alpha threshold,
most commonly 0.05, are interpreted as evidence that the data are
sufficiently inconsistent with the null hypothesis (``significant'').
P-values above that threshold are considered consistent with the null
hypothesis and practically lead to rejecting the alternative hypothesis
(``non-significant''). In (1) above, the reported p-value suggest a
significant result, the p-value derived from the degrees of freedom and
the test statistic suggest a non-significant result. In the following we
refer to these inconsistencies as ``decision inconsistencies''.

The consistency of these values can be automatically assessed if
statistical tests are reported in an unambiguous format. Recently, a
series of studies used such automatic assessments to evaluate the
prevalence of inconsistent statistical reporting in psychology
(\citeproc{ref-bakker2011mis}{Bakker \& Wicherts, 2011},
\citeproc{ref-bakker2014outlier}{2014};
\citeproc{ref-caperos2013consistency}{Caperos \& Pardo, 2013};
\citeproc{ref-claesen2023data}{Claesen et al., 2023};
\citeproc{ref-green2018statcheck}{Green et al., 2018};
\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016};
\citeproc{ref-nuijten2020statcheck}{Nuijten \& Polanin, 2020};
\citeproc{ref-veldkamp2014statistical}{Veldkamp et al., 2014};
\citeproc{ref-wicherts2011willingness}{Wicherts et al., 2011}), medical
sciences (\citeproc{ref-garcia2004incongruence}{Garcı́a-Berthou \&
Alcaraz, 2004}; \citeproc{ref-van2023comparing}{Van Aert et al., 2023}),
psychiatry (\citeproc{ref-berle2007inconsistencies}{Berle \& Starcevic,
2007}), cyber security studies (\citeproc{ref-gross2021fidelity}{Groß,
2021}), technological education research
(\citeproc{ref-buckley2023estimating}{Buckley et al., 2023}), and
experimental philosophy (\citeproc{ref-colombo2018statistical}{Colombo
et al., 2018}). For example, looking at over 250 thousand p-values
published in major psychology journals, Nuijten et al.
(\citeproc{ref-nuijten2016prevalence}{2016}) found that around 50\% of
the articles with statistical results contained at least one
inconsistency and around 13\% contained at least one ``decision
inconsistency''. Other studies report on inconsistency rates between 4\%
and 14\%, with between 10\% and 63\% of articles containing at least one
inconsistency and between 3\% and 21\% decision consistencies.

To assess the prevalence of statistical-reporting inconsistencies in
experimental linguistics, the present paper conceptually replicates
Nuijten et al. (\citeproc{ref-nuijten2016prevalence}{2016}) and assesses
p-values reported in eight experimental linguistic journals published
between 2000 and 2023. We explore whether the rate of inconsistency
differs across journals, whether that rate has changed over the course
of the last 20 years and whether there is evidence for bias in these
statistical-reporting inconsistencies. We discuss the results and offer
concrete recommendations for authors, reviewers, and editors to tackle
this problem.

@@

All quantitative analyses were conducted using R Core Team
(\citeproc{ref-rmanual}{2021}) and the r packages {[}LIST ALL PACKAGES
from 01\_Analysis.R and add to references{]}

@@

We used the R package wtatcheck (Version 1.4.1-beta.2,
\citeproc{ref-nuijten2023statcheck}{Nuijten \& Epskamp, 2023}) to
automatically detect statistical-reporting inconsistencies. Statcheck
works as follows: After converting articles in pdf or html format to
plain text, Statcheck searches for specific strings that correspond to a
NHST result using ``regular expressions''. That way, statcheck can
detect results of t-tests, F-tests, Z-tests, χ2-tests, correlation
tests, and Q tests as long as the test result fulfills three conditions:
(a) the test result is reported completely including the test statistic,
the degrees of freedom (if applicable), and the p-value; (b) the test
result is in the body of the text, i.e.~Statcheck usually misses
reported in tables; and (c) the test result is reported in American
Psychological Association style (\citeproc{ref-APA2020}{American
Psychological Association, 2020}). Given these constraints, Statcheck is
estimated to detect roughly 60\% of all reported NHST results
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). Statcheck
uses the reported test statistic and degrees of freedom to recalculate
the p-value, compares the reported and recalculated p-values and, if
there is a mismatch, flags the test as containing an ``inconsistency.''
The algorithm takes into account that tests might have been performed as
one-tailed by identifying the search strings ``one-tailed,''
``one-sided,'' or ``directional'' in the body of the text. Moreover,
Statcheck considers p = .000 and p \textless{} .000 as inconsistent
because p-values of exactly zero are mathematically impossible and the
APA manual (\citeproc{ref-APA2020}{American Psychological Association,
2020}) advises to report very small p-values as p \textless{} .001.
Validity checks of Statcheck suggest that inter-rater reliability
between manual coding and Statcheck is high, i.e.~0.76 for
inconsistencies and 0.89 for decision inconsistencies
(\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). The
overall accuracy of Statcheck is estimated to be between 96.2\% to
99.9\% (\citeproc{ref-nuijten2017validity}{Nuijten et al., 2017}). We
thus consider Statcheck a valid proxy of the prevalence of statistical
reporting inconsistencies.

@@

Focusing on experimental linguistic research, we used Kobrock and
Roettger (\citeproc{ref-kobrock2023assessing}{2023}) as a point of
departure who list 100 linguistic journals that had at least a hundred
articles published at the time of assessment (2021) and a high ratio of
articles containing the term ``experiment* in title, abstract and/or
keywords. Out of these 100 journals, we selected all journals with at
least 10\% of articles containing the search string''experiment*``. Out
of the remaining 37 journals, we selected those journals that urged APA
formatting either in the main body of the text or specifically regarding
statistics in the author guidelines, resulting in nine remaining
journals. Moreover, to access the articles in .pdf format, the articles
had to be either accessible to us through our library license, or open
access, resulting in a final list of eight journals: Applied
Psycholinguistics (APS), Bilingualism: Language and Cognition (BLC),
Linguistic Approaches to Bilingualism (LAB), Language and Speech (LaS),
Language Learning and Techology (LLT), Journal of Language and Social
Psychology (LSP), Journal of Child Language (JCL), and Studies in Second
Language Acquisition (SLA).

We included only original research articles within the publication years
of 2000-2023, excluding any book reviews, response articles,
commentaries, editorials, corrigenda, errata, advertisements, etc.
Articles from LAB spanned 2011-2023, while the other journals spanned
2000-2023. Statcheck could not parse 157 articles, likely related to
issues with rendering the Chi-Squared symbol being erroneously converted
from .pdf to .txt. This procedure resulted in 5804 research articles
which were submitted to analysis.

@@

All derived data and corresponding R scripts are publicly available
here: LINK.

@@

The results are summarized in Table~\ref{tbl-summary}. Out of 5804
articles, 3059 articles contained statistical tests that statcheck could
assess (53\%), amounting to 39532 assessable p-values. 5166 p-values
were flagged as inconsistent (13.1\%) out of which 655 were considered
decision inconsistencies (1.7\%) (see Table~\ref{tbl-summary}).

@@

{\caption{{Number of eligible articles, assessable articles and results,
inconsistencies and decision inconsistencies across all journals.
(Applied Psycholinguistics (APS), Bilingualism: Language and Cognition
(BLC), Linguistic Approaches to Bilingualism (LAB), Language and Speech
(LaS), Language Learning and Techology (LLT), Journal of Language and
Social Psychology (LSP), Journal of Child Language (JCL), and Studies in
Second Language Acquisition (SLA))}{\label{tbl-summary}}}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0755}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1698}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1887}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1792}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1509}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2358}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Journal
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
eligible articles
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
assessable articles
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
assessible results
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
inconsistencies
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
decision inconsistencies
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
APS & 953 & 690 & 9570 & 1368 & 170 \\
BLC & 964 & 610 & 9093 & 1161 & 120 \\
JCL & 1109 & 529 & 6240 & 750 & 69 \\
LAB & 471 & 133 & 1719 & 234 & 25 \\
LLT & 421 & 111 & 919 & 201 & 61 \\
LSP & 695 & 376 & 4320 & 429 & 60 \\
LaS & 598 & 363 & 4717 & 552 & 86 \\
SLA & 593 & 247 & 2954 & 471 & 64 \\
Total & 5804 & 3059 & 39532 & 5166 & 655 \\
\end{longtable}

@@

\caption{\label{fig-stacked-bar}Proportion of articles containing at
least one inconsistency / decision inconsistency. (Applied
Psycholinguistics (APS), Bilingualism: Language and Cognition (BLC),
Linguistic Approaches to Bilingualism (LAB), Language and Speech (LaS),
Language Learning and Techology (LLT), Journal of Language and Social
Psychology (LSP), Journal of Child Language (JCL), and Studies in Second
Language Acquisition (SLA))}

@@

\caption{\label{fig-year-prop}Proportion of inconsistencies / decision
inconsistencies plotted from 2000 to 2023. Left: rates pooled across
journals; Right: rates per journal. (Applied Psycholinguistics (APS),
Bilingualism: Language and Cognition (BLC), Linguistic Approaches to
Bilingualism (LAB), Language and Speech (LaS), Language Learning and
Techology (LLT), Journal of Language and Social Psychology (LSP),
Journal of Child Language (JCL), and Studies in Second Language
Acquisition (SLA))}

@@

When examining reported against recalculated p-values for all
inconsistencies (see Figure~\ref{fig-scatter}), we can identify certain
spatial patterns (only inconsistencies that are based on `=' comparisons
were included to make them easier to interpret, n = 3779 out of 5166).
First, the center of density of points is located in the bottom left
corner with more p-values reported closer to the standard alpha level of
0.05. This is not surprising, since the quantitative sciences exhibit a
documented publication bias, with hypotheses being more often confirmed
(and thus p \textless{} .05) than not
(\citeproc{ref-franco2014publication}{Franco et al., 2014};
\citeproc{ref-sterling1959publication}{Sterling, 1959}). Second, there
is a linear density of points along the diagonal axis which corresponds
to numerically small inconsistencies, some of which might be related to
simple rounding errors. However, comparing the diagonal to the black
line, which represents a linear model predicting recalculated by
reported p-values, we can see a clear divergence of what is expected if
inconsistencies were equally likely in both directions
(i.e.~recalculated p-values were as likely to be smaller than the
reported p-value as larger). The slope of the regression line is flatter
than the diagonal axis which means that, on average, inconsistencies
have a tendency to exhibit smaller reported p-values than their
recalculated counterparts. Similarly, inconsistencies that report the
p-value as being larger or smaller than a reference value (e.g.~p
\textgreater{} 0.05 or p \textless{} 0.05 ) are not equally prevalent.
There were 4.4\% of inconsistencies with p being reported as larger than
a reference but 7.4\% inconsistencies with p being reported as smaller
than a reference (e.g.~p \textless{} 0.05). So even if we assumed these
inconsistencies were merely typos of the comparison sign,
inconsistencies that erroneously report the p-value to be smaller than a
reference value are more frequent than inconsistencies that erroneously
report the p-value to be larger than a reference value. These biases are
also reflected in decisions inconsistencies. Of all decisions
inconsistencies (n = 655), 71\% represent cases in which a reported
significant result (p \textless{} 0.05) is recalculated as
non-significant (p \textgreater{} 0.05).

@@

@@

The present study found a large amount of statistical reporting
inconsistencies across a sample of 5804 experimental linguistic
articles, containing 3059 assessable p-values. 13.1\% of all p-values
were flagged as inconsistent and 1.7\% were flagged as decision
inconsistencies, i.e.~the reported p-value is on the opposite side of
the alpha threshold than the recalculated p-value. On average, 50\% of
assessable articles contained at least one inconsistency and 13\%
contained at least one decision inconsistency.

The present study can be considered a conceptual replication of previous
studies investigating statistical-reporting inconsistency across
different disciplines (\citeproc{ref-bakker2011mis}{Bakker \& Wicherts,
2011}, \citeproc{ref-bakker2014outlier}{2014};
\citeproc{ref-caperos2013consistency}{Caperos \& Pardo, 2013};
\citeproc{ref-veldkamp2014statistical}{Veldkamp et al., 2014};
\citeproc{ref-wicherts2011willingness}{Wicherts et al., 2011}) and most
recent such assessment using Statcheck
(\citeproc{ref-buckley2023estimating}{Buckley et al., 2023};
\citeproc{ref-colombo2018statistical}{Colombo et al., 2018};
\citeproc{ref-gross2021fidelity}{Groß, 2021};
\citeproc{ref-nuijten2016prevalence}{Nuijten et al., 2016}). Previous
studies report on inconsistency rates between 4\% and 14\%, with between
10\% and 63\% of articles containing at least one inconsistency and
between 3\% and 21\% contain at least one decision consistencies.

Even if the prevalence of these inconsistencies could be largely
attributed to inconsequential typos or rounding errors (an assumption we
cannot test without access to the data), the sheer amount of these
inconsistencies that have made it through peer-review should concern us.
They are human errors. If such a substantial amount of errors is found
in plain site, the question naturally arises as to how many errors
during the data analysis itself remain undetected. If the tip of the
iceberg above water is already so large, how large is the body of the
iceberg underneath the surface?

The present examination assessment did not indicate any noticeable trend
over the last 23 years of publications nor did the data suggest large
differences between journals.
